\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{\numberline{}Acknowledgments} 

\hskip 15pt 
Many people beyond this document's author list have
contributed to the development of dynamics and physics.  
For the dynamics solver
development, we would like to thank Louis Wicker for his assistance with
the Runge-Kutta integration scheme; George Bryan and Jason Knievel
for contributing in the turbulent mixing algorithms and other model filters;
Mark Richardson, Anthony Toigo and Claire Newman for generalizing the
WRF code for global applications. 

\vskip 10pt
For physics contributions, we would like to thank Tom Black, John Brown, Kenneth Campana, 
Fei Chen, Shu-Hua Chen, Ming-Da Chou, Aijun Deng, Mike Ek, Brad Ferrier, Georg Grell, Bill
Hall, Songyou Hong, Zavisa Janjic, Jack Kain, Hiroyuki Kusaka, 
Ruby Leung, Jeong-Ock Lim, Kyo-Sun Lim, Yubao Liu, Chin-Hoh Moeng,
Ken Mitchell, Eli Mlawer, Hugh Morrison, Jon Pleim, Tanya Smirnova, Dave Stauffer, 
Wei-Kuo Tao, Mukul Tewari, Greg Thompson, Guenther Zaengl and others.
We also acknowledge the contribution by Min Chen, Ju-Won Kim, Steven Peckham, Tanya Smirnova,
John Brown and Stan Benjamin for the digital filter initialization development.

\vskip 10pt
The development of WRF-Var represents an international team effort. 
We would like to acknowledge the following people for their
contributions to the WRF-Var system: Yong-Run Guo, Wei Huang, Qingnong
Xiao, Syed Rizvi, Francois Vandenberghe, Mike McAtee, Roy Peck, Wan-Shu Wu,
Dezso Devenyi, Jianfeng Gu, Mi-Seon Lee, Ki-Han Youn, Eunha Lim, Hyun-Cheol Shin,
Shu-Hua Chen, Hui-Chuan Lin, John Bray, Xin Zhang, Zhiquan Liu,
Tom Auligne, Xianyan Zhang, and Yongsheng Chen.

\vskip 10pt
We would like express our appreciation to John Michalakes for developing
the WRF Software Framework.  We would also like thank Jacques Middlecoff, 
Tom Henderson, Al Bourgeois, Matt Bettencourt, Todd Hutchinson, Kent Yang,
and Dan Schaffer for their contributions to the software development. 

\vskip 10pt
Finally, we would like to thank George Bryan,
Jim Bresch, Jian-Wen Bao and John Brown
for their careful review of this document.  We also thank 
Cindy Bruy\`ere for assistance with the graphics.

\vskip 10pt
The Weather Research and Forecasting Model effort
is supported by the National Science Foundation, the Department of
Defense, the National Oceanic and Atmospheric Administration, and the
Federal Aviation Administration.

